Charged pion emission from neutron-rich heavy ion collisions for studies on the symmetry energy

### 2020-08-21 MSU-thesis_estee_MaxFormat.tex
Second revisions to University
Final submit to Uni

\usepackage{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{overpic}
\usepackage{contour}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{rotating}
\usepackage{mathtools}
\usepackage{comment}\renewcommand{\arraystretch}{1.5}
\usepackage[version=3]{mhchem}
\usepackage{newtxtext,newtxmath,booktabs,siunitx}
\usepackage[export]{adjustbox}

@@

@@

Recent astronomical measurements of neutron star mergers have reinvigorated questions about the nature of dense matter in the universe. The densities reached in these stars range from normal nuclear densities $\rho_o$ to upwards of $4\rho_o$ or even $9\rho_o$ in some models, where $\rho_o = \SI{1.7e14}{\gram\per\centi\metre\cubed}$; representing extremely dense, neutrons rich matter. The symmetry energy describes the difference in binding energy between pure neutron matter and symmetric matter with equal number of neutrons and protons. It is the density dependence of the symmetry energy that is related to the outward pressure, which supports the star, preventing it from collapsing under its own gravitation force. For years, theory and experiment  have combined to progressively constrain the Equation of State (EoS) of symmetric matter, and it was only in the last decade that the density dependence of pure neutron matter, the symmetry energy, had been constrained at normal nuclear densities and lower, $\rho < \rho_o$. Still large uncertainties, and discrepancies on preliminary constraints, remain at densities around 2$\rho_o$ which are critical importance to the understanding the properties of neutron stars.

Besides observing neutrons stars directly, the laboratory provides the only way to study dense nuclear matter at different densities and proton-neutron asymmetries. It has been shown that charged pions are produced from nucleon-nucleon collisions through the short-lived intermediate $\Delta$ resonances, in the early, dense parts of Heavy Ion Collisions (HICs). Yet, existing pion experimental data tailored to answering questions about the symmetry energy was lacking and suffers from unknown systematic errors. This prompted the construction of the Samurai Pion-Reconstruction and Ion Tracker Time Projection Chamber (\spirit TPC) which was constructed to measure charged pions produced in neutron-rich HICs with a high efficiency. This Dissertation encompasses the first pion spectra from the most neutron-rich ($\tin{132}{124}$), and neutron-poor ($\tin{108}{112}$) systems measured at \SI{270}{\MeVA} taken at the RIBF facility at RIKEN. The total pion ratio was measured to within 4\% and the pion spectra are measured with a high degree of accuracy, marking the first time pion spectra  -- at energies relevant to the symmetry energy -- have been published.  Also, we will compare the results with the most recent transport models which can simulate pion production. We will highlight some of the successes and hope to motivate a discussion in the theoretical community on how to better reproduce the pion phenomena in neutron-rich HICs, which is important for not only constraining the density dependence of the symmetry energy, but also for understanding the $\Delta$ and pion roles in neutron stars. 
\end{abstract}

@@

I would like to also thank my greater collaboration, starting with the Japanese. Thank you Tadaaki Isobe and Mizuki Kurata-Nishimura, Professor Tetsuya Murakami, and Masanori Kaneko for leading the effort in Japan and seeing this experiment through. Isobe-san, I wanted to thank you especially for the hard work of setting up the DAQ and electronics of the TPC which really preformed flawlessly, considering how many subsystems were merged and the newness of the GET electronics. Mizuki-san, I always enjoyed working with you. You are a hard worker and without you the experimental setup, testing, and analysis could not have finished. Murakami-san, thank you for your support, advice, and wealth of knowledge. I considered you my advisor when I was away from home. Kaneko-san, thank you for working so diligently on the Kyoto array and the analysis. To Hidetada Baba, thank you for working on the electronics of our experiment, DAQ, and for introducing me to the wonders of Japanese Baseball. To all my friends at the University of Tokyo CNS for your friendship, softball, and hospitality. I also wanted to thank our Polish colleagues Jerzy \L ukasik, Piotr Paw\l owski, Krzysztof Pelczar, and Pawe\l \L asko for their hard work in developing the KATANA array and for setting up the trigger in the experiment.

@@

\mainmatter
%%I chose to have each of my chapters in a separate file in a subdirectory called ``chapters''.  If you want to do that, 
%%here's how: - DMA
\include{chapters_altFormat/ch_intro}
\include{chapters_altFormat/ch_experiment}
\include{chapters_altFormat/ch_data_analysis}
\include{chapters_altFormat/ch_data_analysis_2}
\include{chapters_altFormat/ch_results}
\include{chapters_altFormat/ch_summary}
%%%%%%%    APPENDICES    %%%%%%%%%%
%% Uncomment ``\appendix'' for one appendix, ``\appendices'' for more than one, and neither for none.
%%\appendix
\appendices
\include{chapters_altFormat/ch_appendix}
%%If you're not using the \include method, 
%% just type \chapter{<appendix 1 name>}
%% and enter the text of the appendix as you would a chapter.

>>>>

\usepackage{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{overpic}
\usepackage{contour}
\usepackage{color}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{rotating}
%\usepackage{showframe}
\usepackage{mathtools}
\usepackage{comment}\renewcommand{\arraystretch}{1.5}
\usepackage[version=3]{mhchem}
\usepackage{newtxtext,newtxmath,booktabs,siunitx}
\usepackage[export]{adjustbox}

@@

\makeatletter
\patchcmd{\@makechapterhead}{\vspace*{50\p@}}{\vspace*{60pt}}{}{}%
\patchcmd{\@makeschapterhead}{\vspace*{50\p@}}{\vspace*{60pt}}{}{}%
\makeatother

@@

Recent astronomical measurements of neutron star mergers have reinvigorated questions about the nature of dense matter in the universe. The densities reached in these stars range from normal nuclear densities $\rho_o$ to upwards of $4\rho_o$ or even $9\rho_o$ in some models, where $\rho_o = \SI{1.7e14}{\gram\per\centi\metre\cubed}$; representing extremely dense, neutrons rich matter. The symmetry energy describes the difference in binding energy between pure neutron matter and symmetric matter with equal number of neutrons and protons. It is the density dependence of the symmetry energy that is related to the outward pressure, which supports the star, preventing it from collapsing under its own gravitation force.

 For years, theory and experiment  have combined to progressively constrain the Equation of State (EoS) of symmetric matter, and it was only in the last decade that the density dependence of pure neutron matter, the symmetry energy, had been constrained at normal nuclear densities and lower, $\rho < \rho_o$. Still large uncertainties, and discrepancies on preliminary constraints, remain at densities around 2$\rho_o$ which are critical importance to the understanding the properties of neutron stars.

Besides observing neutrons stars directly, the laboratory provides the only way to study dense nuclear matter at different densities and proton-neutron asymmetries. It has been shown that charged pions are produced from nucleon-nucleon collisions through the short-lived intermediate $\Delta$ resonances, in the early, dense parts of Heavy Ion Collisions (HICs). Yet, existing pion experimental data tailored to answering questions about the symmetry energy was lacking and suffers from unknown systematic errors. This prompted the construction of the Samurai Pion-Reconstruction and Ion Tracker Time Projection Chamber (\spirit TPC) which was constructed to measure charged pions produced in neutron-rich HICs with a high efficiency.

This Dissertation encompasses the first pion spectra from the most neutron-rich ($\tin{132}{124}$), and neutron-poor ($\tin{108}{112}$) systems measured at \SI{270}{\MeVA} taken at the RIBF facility at RIKEN. The total pion ratio was measured to within 4\% and the pion spectra are measured with a high degree of accuracy, marking the first time pion spectra  -- at energies relevant to the symmetry energy -- have been published.  Also, we will compare the results with the most recent transport models which can simulate pion production. We will highlight some of the successes and hope to motivate a discussion in the theoretical community on how to better reproduce the pion phenomena in neutron-rich HICs, which is important for not only constraining the density dependence of the symmetry energy, but also for understanding the $\Delta$ and pion roles in neutron stars. 
\end{abstract}

@@

I would like to also thank my greater collaboration, starting with the Japanese. Thank you Tadaaki Isobe and Mizuki Kurata-Nishimura, Professor Tetsuya Murakami, and Masanori Kaneko for leading the effort in Japan and seeing this experiment through. Isobe-san, I wanted to thank you especially for the hard work of setting up the DAQ and electronics of the TPC which really preformed flawlessly, considering how many subsystems were merged and the newness of the GET electronics. Mizuki-san, I always enjoyed working with you. You are a hard worker and without you the experimental setup, testing, and analysis could not have finished. Murakami-san, thank you for your support, advice, and wealth of knowledge. I considered you my advisor when I was away from home. Kaneko-san, thank you for working so diligently on the Kyoto array and the analysis. To Hidetada Baba, thank you for working on the electronics of our experiment, DAQ, and for introducing me to the wonders of Japanese Baseball. To all my friends at the University of Tokyo CNS for your friendship, softball, and hospitality. I also wanted to thank our Polish colleagues Jerzy \L ukasik, Piotr Paw\l owski, Krzysztof Pelczar, and Pawe\l ~\L asko for their hard work in developing the KATANA array and for setting up the trigger in the experiment.

@@

\mainmatter
%%I chose to have each of my chapters in a separate file in a subdirectory called ``chapters''.  If you want to do that, 
%%here's how: - DMA
\include{chapters_altFormat/ch_intro}
\include{chapters_altFormat/ch_experiment}
\include{chapters_altFormat/ch_data_analysis}
\include{chapters_altFormat/ch_data_analysis_2}
\include{chapters_altFormat/ch_results}
\include{chapters_altFormat/ch_summary}
%%%%%%%    APPENDICES    %%%%%%%%%%
%% Uncomment ``\appendix'' for one appendix, ``\appendices'' for more than one, and neither for none.
\appendix
%\appendices
\include{chapters_altFormat/ch_appendix}
%%If you're not using the \include method, 
%% just type \chapter{<appendix 1 name>}
%% and enter the text of the appendix as you would a chapter.fil

### 2020-08-21 chapters_altFormat/ch_data_analysis.tex
Second revisions to University
Final submit to Uni

 \label{fig:mom_sc}
\end{figure}

@@

The final position of the electron along the wire is calculated  $x' = x + dx$. The electron will terminate on the closest anode wire, that is the anode wire that is closest to the shifted z-position $z\textprime = z + dz$, and the final electron z position is then updated to be the same as the anode wire it terminated on. The total drift time of the electron $t$ is calculated as,

@@

\subsection{Pad Response Task}
The total charge of each avalanche is then distributed according to the pad response function (PRF) described in Section~\ref{sec:prf}. The PRF is simulated as the double integral of a 2-dimensional Gaussian. The final output charge on all the pads are the superposition of the PRFs of all drifted electrons. The MC PRF is expressed as,

@@

Now that the pion distribution is flattened, we take bin slices along $p/q$ to determine the particle yield in a certain bin and also estimate the background contribution. The background relevant to  the charged pions comes from the $e^+$ and $e^-$ resulting from the $\pi^0$ Dalitz decay described in Appendix~\ref{appen:dalitz}. Figure~\ref{fig:pimmom_flat} shows the Gaussian fits to the $z_i$ distribution for two different $p/q$ binned cuts around the $\pi^-$ particles, whereas Figure~\ref{fig:pipmom_flat} shows the background $\pi^+$ particles. A Gaussian fit is performed to the pions and to the $e$ spectrum to estimate the background contribution.

>>>>

\caption{Example of before and after the space charge correction comparing the momentum measurement with the BDC vertex included and without it.}
 \label{fig:mom_sc}
\end{figure}

@@

The final position of the electron along the wire is calculated  $x' = x + dx$. The electron will terminate on the closest anode wire, that is the anode wire that is closest to the shifted z-position $z\textprime = z + dz$, and the final electron z position is then updated to be the same as the anode wire it terminated on.

 The total drift time of the electron $t$ is calculated as,

@@

\subsection{Pad Response Task}
The total charge of each avalanche is then distributed according to the pad response function (PRF) described in Section~\ref{sec:prf}. The PRF is simulated as the double integral of a 2-dimensional Gaussian. The final output charge on all the pads are the superposition of the PRFs of all drifted electrons.

 The MC PRF is expressed as,

@@

Now that the pion distribution is flattened, we take bin slices along $p/q$ to determine the particle yield in a certain bin and also estimate the background contribution. The background relevant to  the charged pions comes from the $e^+$ and $e^-$ resulting from the $\pi^0 \rightarrow e^+ e^- \gamma$ Dalitz decay. Figure~\ref{fig:pimmom_flat} shows the Gaussian fits to the $z_i$ distribution for two different $p/q$ binned cuts around the $\pi^-$ particles, whereas Figure~\ref{fig:pipmom_flat} shows the background $\pi^+$ particles. A Gaussian fit is performed to the pions and to the $e$ spectrum to estimate the background contribution.

### 2020-08-21 chapters_altFormat/ch_data_analysis_2.tex
Submit to University
Final submit to Uni

\begin{figure}[!htb]%
    \centering
    \subfloat[ $\tin{108}{112}$ system. ]{{\includegraphics[width=.8\textwidth,valign=c]{angularCut-pim-Sn108.png} \label{fig:pim108angle}}}%
    \qquad
    \subfloat[$\tin{132}{124}$ system. ]{{\includegraphics[width=.8\textwidth,valign=c]{angularCut-pim-Sn132.png} \label{fig:pim132angle}}}%

@@

\begin{figure}[!htb]%
    \centering
    \subfloat[ $\tin{108}{112}$ system.]{{\includegraphics[width=.8\textwidth,valign=c]{angularCut-pip-Sn108.png} \label{fig:pip108angle}}}%
    \qquad
    \subfloat[$\tin{132}{124}$ system.]{{\includegraphics[width=.8\textwidth,valign=c]{angularCut-pip-Sn132.png} \label{fig:pip132angle}}}%
     \caption{Measured $\pi^+$ angular distribution for the two spherical angles $\theta$ and $\phi$.}
        \label{fig:pip}
\end{figure}

@@

\subsection{Correcting Pion Spectra (Efficiency and Acceptance)}
\label{sec:corrSpectra}
To correct for the efficiency, we apply a track-by-track correction, where the efficiency of the i-th track $\epsilon_i$ is retrieved from the efficiency database, from the parameters discussed in Section~\ref{sec:efficiency}. The correction factor $C_i$ is defined as,

@@

\[ \left( \begin{array}{cccc}
1 & 0 & 0 & 0\\
0 & 1 & 0 & 0\\
0 & 0 & \gamma & -\beta \gamma \\
0 & 0 & -\beta \gamma & \gamma
\end{array} \right) \]

@@

\subsection{Pion Statistical Error}

>>>>

\begin{figure}[!htb]%
    \centering
    \subfloat[ $\tin{108}{112}$ system. ]{{\includegraphics[width=.9\textwidth,valign=c]{angularCut-pim-Sn108.png} \label{fig:pim108angle}}}%
    \qquad
    \subfloat[$\tin{132}{124}$ system. ]{{\includegraphics[width=.9\textwidth,valign=c]{angularCut-pim-Sn132.png} \label{fig:pim132angle}}}%

@@

\begin{figure}[!htb]%
    \centering
    \subfloat[ $\tin{108}{112}$ system.]{{\includegraphics[width=.9\textwidth,valign=c]{angularCut-pip-Sn108.png} \label{fig:pip108angle}}}%
    \qquad
    \subfloat[$\tin{132}{124}$ system.]{{\includegraphics[width=.9\textwidth,valign=c]{angularCut-pip-Sn132.png} \label{fig:pip132angle}}}%
     \caption{Measured $\pi^+$ angular distribution for the two spherical angles $\theta$ and $\phi$.}
        \label{fig:pip}
\end{figure}

@@

\section{Correcting Pion Spectra (Efficiency and Acceptance)}
\label{sec:corrSpectra}
To correct for the efficiency, we apply a track-by-track correction, where the efficiency of the i-th track $\epsilon_i$ is retrieved from the efficiency database, from the parameters discussed in Section~\ref{sec:efficiency}. The correction factor $C_i$ is defined as,

@@

\begin{equation}
A = \begin{pmatrix}
1 & 0 & 0 & 0\\[-25pt]
0 & 1 & 0 & 0\\[-25pt]
0 & 0 & \gamma & -\beta \gamma\\[-25pt]
0 & 0 & -\beta \gamma & \gamma
\end{pmatrix},
\end{equation}

@@

\section{Pion Statistical Error}
